# Contrasts
```{r, echo=FALSE}
# Unattach any packages that happen to already be loaded. In general this is unecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}

# Set my default chunk options 
knitr::opts_chunk$set( fig.height=3 )
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)   # ggplot2, dplyr, tidyr
library(tidymodels)  # for broom functions
library(emmeans)    # for emmeans()
# library(multcomp)   # for glht()  - multcomp fights with dplyr, :(
```

<!-- ### Learning Outcomes{-} -->

## Introduction {#Contrasts_Introduction}
We have written our model as $\boldsymbol{y} = \boldsymbol{X\beta}+\boldsymbol{\epsilon}$ and often we are interested in some linear function of the $\boldsymbol{\hat{\beta}}$. Some examples include the model predictions $\hat{y}_i = \boldsymbol{X}_{i\cdot} \boldsymbol{\beta}$ where $\boldsymbol{X}_{i\cdot}$ is the $i^{th}$ row of the $\boldsymbol{X}$ matrix. Other examples include differences in group means in a one-way ANOVA or differences in predicted values $\hat{y}_i - \hat{y}_j$.

All of these can can be written as $\boldsymbol{c}'\boldsymbol{\beta}$ for some vector $\boldsymbol{c}$.

We often are interested in estimating a function of the parameters $\boldsymbol{\beta}$. For example in the offset representation of the ANOVA model with 3 groups we have
$$y_{ij}=\mu+\tau_{i}+\epsilon_{ij}$$
where 
$$
\boldsymbol{\beta}=\left[\mu\;\tau_{2}\;\tau_{3}\right]^{T}
$$
and $\mu$ is the mean of the control group, group one is the control group and thus $\tau_{1}=0$, and $\tau_{2}$ and $\tau_{3}$ are the offsets of group two and three from the control group. In this representation, the mean of group two is $\mu+\tau_{2}$ and is estimated with $\hat{\mu} + \hat{\tau}_2$.


A contrast is a linear combinations of elements of $\boldsymbol{\hat{\beta}}$, which is a fancy way of saying that it is a function of the elements of $\boldsymbol{\hat{\beta}}$
  where the elements can be added, subtracted, or multiplied by constants. In particular, the contrast can be represented by the vector $\boldsymbol{c}$ such that the function we are interested in is $\boldsymbol{c}^{T}\boldsymbol{\hat{\beta}}$.

In the ANOVA case with $k=3$ where we have the offset representation, I might be interested in the mean of group 2, which could be written as 
$$
\hat{\mu}_2 = \hat{\mu}+\hat{\tau}_{2}=\underset{\boldsymbol{c}^{T}}{\underbrace{\left[\begin{array}{ccc}
1 & 1 & 0\end{array}\right]}}\cdot\underset{\hat{\boldsymbol{\beta}}}{\underbrace{\left[\begin{array}{c}
\hat{\mu}\\
\hat{\tau_{2}}\\
\hat{\tau_{3}}
\end{array}\right]}}
$$

Similarly in the simple regression case, I will be interested in the height of the regression line at $x_0$. This height can be written as  
$$
\hat{y}_0 =\hat{\beta}_{0}+\hat{\beta}_{1}x_{0}=\underset{\boldsymbol{c}^{T}}{\underbrace{\left[\begin{array}{cc}
1 & x_{0}\end{array}\right]}}\cdot\underset{\hat{\boldsymbol{\beta}}}{\underbrace{\left[\begin{array}{c}
\hat{\beta}_{0}\\
\hat{\beta}_{1}
\end{array}\right]}}
$$
In this manner, we could think of calculating all of the predicted values $\hat{y}_i$ as just the result of the contrast $\boldsymbol{X} \hat{\boldsymbol{\beta}}$ where our design matrix $\boldsymbol{X}$ takes the role of the contrasts.


## Estimate and variance

One of the properties of maximum likelihood estimator (MLEs), is that they are invariant under transformations. Meaning that since $\hat{\boldsymbol{\beta}}$ is the MLE of $\boldsymbol{\beta}$, then $\boldsymbol{c}^{T}\hat{\boldsymbol{\beta}}$ is the MLE of $\boldsymbol{c}^{T}\boldsymbol{\beta}$. The only thing we need to perform hypotheses tests and create confidence intervals is an estimate of the variance of $\boldsymbol{c}^{T}\hat{\boldsymbol{\beta}}$. 

Because we know the variance of $\hat{\boldsymbol{\beta}}$ is
$$
Var\left(\hat{\boldsymbol{\beta}}\right)=\sigma^{2}\left(\boldsymbol{X}^{T}\boldsymbol{X}\right)^{-1}
$$
and because $\boldsymbol{c}$ is a constant, then 
$$
Var\left(\boldsymbol{c}^{T}\hat{\boldsymbol{\beta}}\right)=\sigma^{2}\boldsymbol{c}^{T}\left(\boldsymbol{X}^{T}\boldsymbol{X}\right)^{-1}\boldsymbol{c}
$$
and the standard error is found by plugging in our estimate of $\sigma^{2}$ and taking the square root.
$$
StdErr\left(\boldsymbol{c}^{T}\hat{\boldsymbol{\beta}}\right)	=	\sqrt{\hat{\sigma}^{2}\boldsymbol{c}^{T}\left(\boldsymbol{X}^{T}\boldsymbol{X}\right)^{-1}\boldsymbol{c}}
	=	\hat{\sigma}\sqrt{\boldsymbol{c}^{T}\left(\boldsymbol{X}^{T}\boldsymbol{X}\right)^{-1}\boldsymbol{c}}
$$ 

As usual, we can now calculate confidence intervals for $\boldsymbol{c}^{T}\hat{\boldsymbol{\beta}}$ using the usual formula
$$Est	\pm	t_{n-p}^{1-\alpha/2}\;StdErr\left(\,Est\,\right)$$
$$
\boldsymbol{c}^{T}\hat{\boldsymbol{\beta}}	\pm	t_{n-p}^{1-\alpha/2}\;\hat{\sigma}\sqrt{\boldsymbol{c}^{T}\left(\boldsymbol{X}^{T}\boldsymbol{X}\right)^{-1}\boldsymbol{c}}
$$ 


Recall the hostility example which was an ANOVA with three groups with the data

| Method   | Test Scores                        |
|:--------:|:-----------------------------------|
|   1      | 96 79 91 85 83 91 82 87            |
|   2      | 77 76 74 73 78 71 80               |
|   3      | 66 73 69 66 77 73 71 70 74         |

We have analyzed this data using both the cell means model and the offset and we will demonstrate how to calculate the group means from the offset representation. Thus we are interested in estimating $\mu+\tau_{2}$ and $\mu+\tau_{3}$.  I am also interested in estimating the difference between treatment 2 and 3 and will therefore be interested in estimating $\tau_{2} - \tau_{3}$.

```{r}
data <- data.frame(
  y =  c(96,79,91,85,83,91,82,87,
         77,76,74,73,78,71,80,
         66,73,69,66,77,73,71,70,74),
  group = factor(c( rep('Group1',8), rep('Group2',7),rep('Group3',9) ))
)

ggplot(data, aes(x=group, y=y)) + geom_boxplot()
```

We can fit the offset model and obtain the design matrix and estimate of $\hat{\sigma}$ via the following code. 
```{r}
m <- lm(y ~ group, data=data)          # Fit the ANOVA model (offset representation)
tidy(m)                                # Show me beta.hat
X <- model.matrix(m)                   # obtains the design matrix
sigma.hat <- glance(m) %>% pull(sigma) # grab sigma.hat 
beta.hat <- tidy(m) %>% pull(estimate) 
XtX.inv <- solve( t(X) %*% X )
```

Now we calculate 
```{r}
contr <- c(1,1,0)  # define my contrast
ctb <- t(contr) %*% beta.hat
std.err <- sigma.hat * sqrt( t(contr) %*% XtX.inv %*% contr )

data.frame(Estimate=ctb, StdErr=std.err)
```


and notice this is the exact same estimate and standard error we got for group two when we fit the cell means model.
```{r}
CellMeansModel <- lm(y ~ group - 1, data=data)
CellMeansModel %>% tidy()
```


## Estimating contrasts using `glht()`

Instead of us doing all the matrix calculations ourselves, all we really need is to is specify the row vector $\boldsymbol{c}^{T}$. The function that will do the rest of the calculations is the generalized linear hypothesis test function `glht()` that can be found in the multiple comparisons package `multcomp`. The p-values will be adjusted to correct for testing multiple hypothesis, so there may be slight differences compared to the p-value seen in just the regular summary table.

### 1-way ANOVA {#Contrasts_glht_OneWayAnova}

We will again use the Hostility data set and demonstrate how to calculate the point estimates, standard errors and confidence intervals for the group means given a model fit using the offset representation. 

```{r}
m <- lm(y ~ group, data=data)
m %>% tidy()
```

We will now define a row vector (and it needs to be a matrix or else `glht()` will throw an error. First we note that the simple contrast $\boldsymbol{c}^{T}=\left[1\;0\;0\right]$ just grabs the first coefficient and gives us the same estimate and standard error as the summary did.
```{r, message=FALSE, warning=FALSE}
contr <- rbind("Intercept"=c(1,0,0))      # 1x3 matrix with row named "Intercept"
test <- multcomp::glht(m, linfct=contr)   # the linear function to be tested is contr
summary(test)
```

Next we calculate the estimate of all the group means $\mu$, $\mu+\tau_{2}$ and $\mu+\tau_{3}$ and the difference between group 2 and 3. Notice I can specify more than one contrast at a time.
```{r}
contr <- rbind("Mean of Group 1"=c(1,0,0),
               "Mean of Group 2"=c(1,1,0),
               "Mean of Group 3"=c(1,0,1),
               "Diff G2-G3"     =c(0,1,-1)) 
test <- multcomp::glht(m, linfct=contr)
summary(test)
```

Finally we calculate confidence intervals in the usual manner using the `confint()` function.
```{r}
confint(test, level=0.95)
```




## Using `emmeans` Package {#Contrasts_emmeans}
Specifying the contrasts by hand is extremely difficult to do correctly and instead we would prefer to specify the contrasts using language like "create all possible pairwise contrasts" where each pair is just a subtraction.  The R-package `emmeans` tries to simply the creation of common contrasts. 

To show how to use the `emmeans` package, we'll consider a bunch of common models and show how to address common statistical questions for each.

### Simple Regression {#Contrasts_SimpleRegression}
There is a dataset built into R named `trees` which describes a set of $n=31$ cherry trees and the goal is to predict the volume of timber produced by each tree just using the tree girth a 4.5 feet above the ground.

```{r}
data(trees)
model <- lm( Volume ~ Girth, data=trees )
trees <- trees %>%
  dplyr::select( -matches('fit'), -matches('lwr'), -matches('upr') ) %>%
  cbind( predict(model, interval='conf'))
ggplot(trees, aes(x=Girth, y=Volume)) +
  geom_point() +
  geom_ribbon( aes(ymin=lwr, ymax=upr), alpha=.3 ) +
  geom_line( aes(y=fit) )
```

Using the `summary()` function, we can test hypotheses about if
the y-intercept or slope could be equal to zero, but we might be interested in confidence intervals for the regression line at girth values of 10 and 12.

```{r}
# We could find the regression line heights and CI using 
# either predict() or emmeans()
predict(model, newdata=data.frame(Girth=c(10,12)), interval='conf' )
emmeans(model, specs = ~Girth, at=list(Girth=c(10,12)) )  
```

The `emmeans()` function requires us to specify the grid of reference points we are interested as well as which variable or variables we wish to separate out.  In the simple regression case, the `specs` argument is just the single covariate.

We might next ask if the difference in volume between a tree with 10 inch girth is statistically different than a tree with 12 inch girth?  In other words, we want to test
$$ H_0:\; (\beta_0 + \beta_1\cdot10 ) - (\beta_0 + \beta_1 \cdot 12) = 0$$ 
$$ H_a:\; (\beta_0 + \beta_1\cdot10 ) - (\beta_0 + \beta_1 \cdot 12) \ne 0$$ 

In this case, we want to look at all possible pairwise differences between the predicted values at $10$ and $12$. (The *rev* part just reverses the order in which we do the subtraction.)

```{r}
emmeans(model, specs = revpairwise~Girth,
        at=list(Girth=c(10,12)) ) 
```

Notice that if I was interested in 3 points, we would get all of the differences.

```{r}
emmeans(model, specs = pairwise~Girth,
        at=list(Girth=c(10,11,12)) ) 
```

In this very simple case, the slope parameter is easily available as a parameter value, but we could use the `emtrends()` function to obtain the slope.  

```{r}
emtrends( model, ~Girth, 'Girth' )
```
This output is a bit mysterious because because of the `13.248` component. What has happened is that `emtrends` is telling us the slope of the line at a particular point on the x-axis (the mean of all the girth values). While this doesn't matter in this example, because the slope is the same for all values of girth, if we had fit a quadratic model, it would not.

```{r}
model <- lm( Volume ~ poly(Girth, 2), data=trees )      # Girth + Girth^2
trees <- trees %>% 
  dplyr::select(Volume, Girth) %>% 
  cbind(predict(model, interval='conf'))

ggplot(trees, aes(x=Girth, y=Volume)) +
  geom_point() +
  geom_ribbon( aes(ymin=lwr, ymax=upr), alpha=.3 ) +
  geom_line( aes(y=fit) )

emtrends( model, ~ poly(Girth,2), 'Girth',
          at=list(Girth=c(10,11,12)) )
```


### 1-way ANOVA {#Contrasts_OneWayAnova}
To consider the pairwise contrasts between different levels we will consider the college student hostility data again. A clinical psychologist wished to compare three methods for reducing hostility levels in university students, and used a certain test (HLT) to measure the degree of hostility. A high score on the test indicated great hostility. The psychologist used $24$ students who obtained high and nearly equal scores in the experiment. Eight subjects were selected at random from among the $24$ problem cases and were treated with method 1, seven of the remaining $16$ students were selected at random and treated with method 2 while the remaining nine students were treated with method 3. All treatments were continued for a one-semester period. Each student was given the HLT test at the end of the semester, with the results show in the following table.

```{r}
Hostility <- data.frame(
  HLT = c(96,79,91,85,83,91,82,87,
          77,76,74,73,78,71,80,
          66,73,69,66,77,73,71,70,74),
  Method = c( rep('M1',8), rep('M2',7), rep('M3',9) ) )
```

```{r}
ggplot(Hostility, aes(x=Method, y=HLT)) +
  geom_boxplot()
```

To use the `emmeans()`, we again will use the pairwise command where we specify that we want all the pairwise contrasts between Method levels.

```{r}
model <- lm( HLT ~ Method, data=Hostility )
emmeans(model, specs = pairwise~Method, 
        infer=c(TRUE,TRUE) )     # Print out Confidence Intervals and Pvalues
```


By default, `emmeans()` and `emtrends()` don't provide p-values for testing if the effect could possibly be equal to zero. To get those, you'll need to set the inference option to return both confidence intervals and p-values using the `infer=c(TRUE,TRUE)`, where the two logical values control the inclusion of the confidence interval and p-value, respectively.

## Exercises {#Exercises_Contrasts}
1. The American Community Survey is on ongoing survey being conducted monthly by the US Census Bureau and the package `Lock5Data` has a dataset called `EmployedACS` that has 431 randomly selected anonymous US residents from this survey.
    ```{r}
    data('EmployedACS', package='Lock5Data')
    ?Lock5Data::EmployedACS
    ```
    a) Create a boxplot of the respondents' `Race` and `Income`.
    b) Fit a 1-way ANOVA on these data.
    c) Use `multcomp::glht()` to calculate all the pairwise differences between the race categories.
    d) Use `emmeans()` to calculate all the pairwise differences between the race categories. 

2. We will examine a data set from Ashton et al. (2007) that relates the length of a tortoise’s carapace to the number of eggs laid in a clutch. The data are
    ```{r}
    Eggs <- data.frame(
      carapace    = c(284,290,290,290,298,299,302,306,306,
                      309,310,311,317,317,320,323,334,334),
      clutch.size = c(3,2,7,7,11,12,10,8,8,
                      9,10,13,7,9,6,13,2,8)) 
    ```
    a) Plot the data with carapace as the explanatory variable and clutch size as the response.
    b) We want to fit the model
    $$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \epsilon_i \textrm{ where } \epsilon_i \stackrel{iid}{\sim} N(0, \sigma^2)$$
    To fit this model, we need the design matrix with a column of ones, a column of $x_i$ values and a column of $x_i^2$ values. To fit this model, we could create a new column of $x_i^2$ values, or do it in the formula.
        ```{r}
        Eggs <- Eggs %>% mutate( carapace.2 = carapace^2 )
        
        model <- lm(clutch.size ~ 1 + carapace + carapace.2, data=Eggs)   # one version
        model <- lm(clutch.size ~ 1 + carapace + I(carapace^2), data=Eggs) # Do squaring inside formula
        model <- lm(clutch.size ~ 1 + poly(carapace, degree=2, raw=TRUE), data=Eggs) # using polynomial function
        ```
    c) Use the `predict()` function to calculate the regression predictions and add those predictions to the `Eggs` dataset.
    d) Graph the data with the regression curve.
    e) Use the `emmeans()` function to estimate the clutch sizes for tortoise's with carapace sizes of 300 and 320. Provide a confidence interval for the estimates and a p-value for the hypothesis test that the value could be zero.
    f) Use the `emmeans()` function to estimate the difference between clutch size prediction for tortoise's with carapace sizes of 300 and 320. Provide a confidence interval for the estimates and a p-value for the hypothesis test that the difference could be zero.
    g) Use the `emtrends()` function to find the estimated slope at carapace sizes of 300 and 320. Provide a confidence interval for the estimates and a p-value for the hypothesis test that the value could be zero.
    h) Use the `emtrends()` function to estimate the differences between slopes at carapace sizes 300 and 320. Provide a confidence interval for the estimates and a p-value for the hypothesis test that the difference could be zero.
    



